---
title: Impact evaluation of storm and severe weather events
author: "Martin Halwachs"
date: "31 Juli 2016"
output:
  html_document:
    toc: TRUE
---

# Synopsis
The storm and severe weather data from 1950 to November 2011 was evaluated for their impact. The weather data was roughly grouped into 48 event types. The impact on human health is compiled from the number of fatalities and injuries of each single event. The sum of injuries and fatalities indicates that tornados, (exzessive) heat, thunderstorms, lightning and (flash) floods are the most dangerous events. These events show up again, when calculating the sum of proprietary and crop damage in USD and breaking down to the ten most expensive events. 
**Ten sentences summarizing the result.**

# Used R Libraries
```{r loading libraries, results="hide", message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(knitr)
```

# Data Processing
First the data is loaded from the web using a temporary file. Note that *.bz2 files can be read directly without unzipping.
```{r download data, cache=TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
tempFile <- tempfile()
download.file(fileUrl, tempFile)
data <- read.csv(tempFile) #bz2 can be read directly without unzipping
```

## Tidy Data
The dimension and column headers of the imported data are:
```{r show all column names}
dim(data)
names(data)
```

To reduce memory overhead less important columns are removed
```{r less columns}
data <- select(data, c(BGN_DATE,EVTYPE,END_DATE,FATALITIES:CROPDMGEXP))
```

### Explanation of relevant colum headers
The [documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) does not show the column header names. Therefore the following list shall provide necessary information about the column names. The section in the documentation is statet, if any information on this column was found. 

(@) **BGN_DATE** - begin date of event (not explicitly stated in the documentation)  
These dates are formated properly, while ignoring the included timestamp:
```{r formating begin date}
dateformat <- "%m/%d/%Y"
data$BGN_DATE <- as.Date(data$BGN_DATE, dateformat)
```

(@) **END_DATE** - end date of event (not explicitly stated in the documentation)  
These dates are formated properly, while ignoring the included timestamp (see BGN_DATE for the date format):
```{r formating end date}
data$END_DATE <- as.Date(data$END_DATE, dateformat)
```

(@) **FATALITIES** - number of fatalities

(@) **INJURIES** - number of injured humans

(@) **PROPDMG** - Property damage in actual dollar amount
According to Chapter 2.7: "*Property damage estimates should be entered as actual dollar amounts, if a reasonably accurate estimate from an insurance company or other qualified individual is available.* [...] *Typically, damage refers to damage inflicted to private property (structures, objects, vegetation) as well as public infrastructure and facilities. Specific breakdowns should be stated in the event narrative (refer to Section 2.9), if possible. The number of structures with minor or moderate damage should be indicated, as well as the number of buildings destroyed.* [...]"

(@) **PROPDMGEXP** - Property damage exponent  
Various property damages can be looked up in Appendix B. According to Chapter 2.7: "[...]*"Estimates should be rounded to three significant digits, followed by an alphabetical character signifying the magnitude of the number, i.e., 1.55B for $1,550,000,000. Alphabetical characters used to signify magnitude include 'K' for thousands, 'M' for millions, and 'B' for billions.*[...]", this column provides information on the exponent for the property damage value. The following shows the found indicators with their count:
```{r property damage exponents}
table(data$PROPDMGEXP)
```
**Important notice:** As only 'K', 'M' and 'B' were cited and the empty string seems to stand for no exponent, all others will be neglected, treating them as improper entries (though 'h' or 'H' may stand for hundreds). As 'milli' for 'm' makes no sense and 'k' for 'kilo' was already stated as 'K' above, the three letters will be tretated case-insensitive!
The following table sums up the treatment of this column:  

| indicator | property.damage |
|-----------|-----------------|
| k,K       | PROPDMG\*10E3   |
| m,M       | PROPDMG\*10E6   |
| b,B       | PROPDMB\*10E9   |

The property damage is adjusted according to the above exponents. Finally this column is dismissed from further evaulations.
```{r get accurate property damage}
columns <- data$PROPDMGEXP %in% c("k","K")
with(data, PROPDMG[columns] <- PROPDMG[columns]*1E3)

columns <- data$PROPDMGEXP %in% c("m","M")
with(data, PROPDMG[columns] <- PROPDMG[columns]*1E6)

columns <- data$PROPDMGEXP %in% c("b","B")
with (data, PROPDMG[columns] <- PROPDMG[columns]*1E9)

columns <- !(data$PROPDMGEXP %in% c("","k","K","m","M","b","B"))
data$PROPDMG[columns] <- 0
data <- select(data, -PROPDMGEXP)
```

(@) **CROPDMG** - Crop damage (not further )  
According to Appendix B: "[...]*Crop damage [Crop value/acre]x [#acres]*[...]"

(@) **CROPDMGEXP** - Crop damage exponent  
seems to be the same as the PROPDMGEXP. The following shows the various indicators and their count:
```{r crop damage exponents}
table(data$CROPDMGEXP)
```

In accordance with PROPDMGEXP, the exponents are applied. The costs for crop and property damage are summed and the separate columns are dismissed from further evaluations. 
```{r get accurate crop damage}
columns <- data$CROPDMGEXP %in% c("k","K")
with(data, CROPDMG[columns] <- CROPDMG[columns]*1E3)

columns <- data$CROPDMGEXP %in% c("m","M")
with(data, CROPDMG[columns] <- CROPDMG[columns]*1E6)

columns <- data$CROPDMGEXP %in% c("b","B")
with (data, CROPDMG[columns] <- CROPDMG[columns]*1E9)

columns <- !(data$CROPDMGEXP %in% c("","k","K","m","M","b","B"))
data$CROPDMG[columns] <- 0

data <- mutate(data, COSTS=CROPDMG+PROPDMG)
data <- select(data, -c(PROPDMG, CROPDMG, CROPDMGEXP))
```

(@) **EVTYPE** - Envent type  
Chapter 2.1.1 Table 1, lists several types of Storm Data events. The following list shows some sample of the **`r length(unique(data$EVTYPE))`** different raw imported event types
```{r show data events}
#set.seed(123)
set.seed(0)
sort(sample(unique(data$EVTYPE),45))
```

This small selection already shows that the list of event types is not well formated (e.g. 'Urban flood' and 'Urban Flooding'), includes several summaries and could be reduced to a more compact list of events. This will be achieved in the next steps

First all elements are written lower case and the summaries are removed,
```{r lower case and remove summaries}
data$EVTYPE <- as.factor(str_to_lower(data$EVTYPE))
summarydata <- grepl("summary", data$EVTYPE)
data <- data[!summarydata,]
```
Next the steps is removing unnecessary spaces and wrong signes:
```{r eliminate sign and whitespace errors}
data$EVTYPE <- as.factor(gsub("^[ ]+","",data$EVTYPE))    #whitespace at begin
data$EVTYPE <- as.factor(gsub("[ ]+$","",data$EVTYPE))    #whitespace at end
data$EVTYPE <- as.factor(gsub("[ ]{2,}"," ",data$EVTYPE)) #(multiple whitespaces)
data$EVTYPE <- as.factor(gsub("[\\&;-]{1,3}","/",data$EVTYPE))      #signs instead of slash
data$EVTYPE <- as.factor(gsub(" /","/",data$EVTYPE))      #whitespace before slash
data$EVTYPE <- as.factor(gsub("/ ","/",data$EVTYPE))      #whitespace after slash
data$EVTYPE <- as.factor(gsub("[\\./]([ ]){0,10}$","",data$EVTYPE)) #signs at string end
```

Finally the various event types are reduced to a rough grouping, while generating the group "other" for non-assignable entries.
```{r reduce event types, cache=TRUE}
data$EVTYPE <- as.factor(gsub("^aval.*","avalanche",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(blizzard[ /$]|blow.*snow).*","blizzard",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(co?a?sta?l|beach).*","coastal flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(cold|cool|low temperature$).*","cold/wind chill",data$EVTYPE))
data$EVTYPE <- as.factor(gsub(".*wind ?chill.*","cold/wind chill",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(((dry|wet) )?mi[cr]+oburst).*","thunderstorm wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^dust dev[ie]l.*","dust devil",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^dust ?storm.*","dust storm",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(mild|dr[yo]u?g?h?t?).*","drought",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(ex(cess|trem)|prolong|record|un(usual|season)|very|abnorm|high|severe).*(dry|(low.*rain)).*","drought",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(warm|hot|heat).*","heat",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(ex(cess|trem)|prolong|hyper|record|un(usual|season)|very|abnorm|high|severe).*(heat|warm|hot|tempera|therm|high).*","extreme heat",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(ex(cess|trem)|prolong|record|un(usual|season)|very|abnorm|high|severe).*(cold|cool|chil|low).*","extreme cold/wind chill",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(hypotherm).*","extreme cold/wind chill",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(ex(cess|trem)|prolong|record|un(usual|season)|very|abnorm|high|severe).*(rain|precip).*","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^low temp.*record$","e",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^marine (thunder|tstm).*","marine thunderstorm wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(flash|flood).*(flooo?d|flash).*","flash flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(drift|(mud|land).*slide).*","debris flow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(river|flood|drowning|urban|(small )?stream|street).*","flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(major|locall?y?) (flood|heavy rain|flash flood)$","\\2",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^gust(nado|y).*","thunderstorm wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^hail[ /$].*","hail",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(heavy|record).*snow.*","heavy snow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^he?a?vy.*(rain|shower|precip).*","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^rain.*","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub(".*(high tide|surf).*","high surf",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^high.*wind.*","high wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^hurri.*","hurricane (typhoon)",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^ice.*fog.*","freezing fog",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^ice.*","ice storm",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^lig[hn]tn?ing.*","lightning",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^sleet.*","sleet",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^snow.*","heavy snow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^((severe )?(drowburst|th?un?d?er?|tstm)).*","thunderstorm wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^tropical storm.*","tropical strom",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^volc.*","volcanic ash",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^torn[ad]+o.*","tornado",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^typhoon.*","hurricane (typhoon)",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(grass|forest|brush|wild).*","wildfire",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^way?ter ?spout.*","waterspout",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^rip current.*","rip current",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(strong )?wi?nd.*","strong wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^storm.*","storm surge/tide",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^winter storm.*","winter storm",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^wint[ery ]+(weather|mix).*","winter weather",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(frost|freezing|freeze) [^f][^o][^g].*","frost/freeze",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(freeze|frost)$","frost/freeze",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^driest*","drought",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^downburst$","thunderstorm wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^fog","dense fog",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^vog$","volcanic ash",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^hail.*","hail",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^abnormally wet$","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^wet weather$","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^accumulated snowfall$","heavy snow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^below normal precipitation$","drought",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^black ice$","frost/freeze",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^blowing dust$","dust storm",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^breakup flooding$","flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^dam.*","flash flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(deep|small) hail$","hail",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(patchy )?dense fog.*","dense fog",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^downburst.*","thunderstorm wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^early (frost|snow|freeze).*","winter weather",data$EVTYPE))
data$EVTYPE <- as.factor(gsub(".*cstl flood.*","coastal flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^excessive snow$","heavy snow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub(".*wetness.*","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^extended cold$","extreme cold/wind chill",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^extremely wet$","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^falling snow/ice$","hail",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^first.*","winter weather",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^glaze.*","frost/freeze",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^funnel.*","funnel cloud",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^gradient winds?$","rip current",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^ground blizzard$","blizzard",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(hard|agricultural) freeze$","frost/freeze",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(heavy|high) swells$","high surf",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^high( water| waves|way flooding)$","flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(icy roads|patchy ice)$","frost/freeze",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^lack of snow$","heat",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^lake flood$","lakeshore flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^lake[ /]effect snow$","lake-effect snow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^landspout$","dust devil",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^late.*","heat",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^light .*","winter weather",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^minor flood.*","flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^mixed precip.*","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(moderate|mountain) snow.*","winter weather",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^near record snow$","heavy snow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^rapidly rising water$","flash flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^remnants of floyd$","hurricane (typhoon)",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^rock slide$","debris flow",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(rogue wave|rough seas)$","marine strong wind",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^rural flood$","flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^sml stream fld$","flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^smoke$","dense smoke",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^temperature record$","extreme heat",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^extreme heat$","excessive heat",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^tidal flood.*","flood",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^torrential rain.*","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^unseasonably wet$","heavy rain",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^unusually late snow","heat",data$EVTYPE))
data$EVTYPE <- as.factor(gsub(".*wall cloud.*","funnel cloud",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^whirlwind$","tornado",data$EVTYPE))

data$EVTYPE <- as.factor(gsub("^apache county$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^blow/out.*","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^excessive$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^early rain.*","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^high$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(heavy|high) (mix|seas)$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^landslump$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^marine (accident|mishap)$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^metro storm.*","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^normal precipitation$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^northern lights$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^red flag.*","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^saharan dust$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^seasonal snowfall$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^severe turbulence$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^southeast$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^wake low wind$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^wet snow$","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub(".*(month|year|day).*","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^(non|no severe).*","other",data$EVTYPE))
data$EVTYPE <- as.factor(gsub("^[\\?e]$","other",data$EVTYPE))
```

This brings up the final list of event types of **`r length(unique(data$EVTYPE))`** different entries:
```{r final event type list}
sort(unique(data$EVTYPE))
```

For better addressing, the column headers are reformatted accordingly.
```{r make names}
names(data) <- make.names(tolower(names(data)))
```


## Checking Data
The tidied data has a size of `r dim(data)`, first and last lines look like:
```{r check data lines}
head(data)
tail(data)
```

The structure is as follows.
```{r check data structure}
str(data)
```

# Results
From the tidy data set of the previous sections, several evaluations will be done to demonstrate the impact of the `r length(levels(data$EVTYPE))` events listed.

## Harmfulness to population health
This first chapter addresses the question, how harmful each event types is. Therefore fatalities and injuries are summed up across the entire united states for each single event.
```{r fatality injury preparation}
amounts <- with(data, tapply(fatalities, evtype, sum))
healthdata <- data.frame(variable=as.factor("fatalities"), amount=as.vector(amounts), event=as.factor(names(amounts)))
amounts <- with(data, tapply(injuries, evtype, sum))
healthdata <- rbind(healthdata, data.frame(variable=as.factor("injuries"), amount=as.vector(amounts), event=as.factor(names(amounts))))

healthdata[sample(nrow(healthdata), 20),]
```

All entries without injuries or fatalities are dismissed.
```{r dismiss zeros}
healthdata <- filter(healthdata, amount>0)
```

As tornodas have by far the highest amount of injuries,
```{r tornado counts}
tornadoInjuries <- filter(healthdata, variable=='injuries', event=='tornado')$amount
tornadoFatalities <- filter(healthdata, variable=='fatalities', event=='tornado')$amount
```
with **`r print(tornadoInjuries)`** injuries and **`r print(tornadoFatalities)`** fatalities, this event is removed for the following plot to show better relations.
```{r dismiss tornados}
healthdata.reduced <- healthdata[!healthdata$event=="tornado",]
```

The following plot shows the amount of fatalities and injuries for each event type.
```{r fatality and injury plot}
p <- ggplot(healthdata.reduced, aes(x=reorder(event, desc(event)),y=amount,fill=variable))
p + geom_bar(stat='identity', position='dodge') + coord_flip() + labs(x="event types", fill="")
```

The graphic indicates several very dramatic events. The 10 most harmful events for injuries and fatalities are found and divided by the number of years recorded to calculate the impact per year
```{r top fatality and injury}
topFatalities <- head(arrange(filter(healthdata, variable=='fatalities'), desc(amount)),10)
topFatalities <- mutate(topFatalities, fatalities.per.year=amount/(2011-1950))
topInjuries <- head(arrange(filter(healthdata, variable=='injuries'), desc(amount)),10)
topInjuries <- mutate(topInjuries, injuries.per.year=amount/(2011-1950))
```

The following tables shows the highest impact events with most occuring injuries and fatalities per year:
```{r fatality and injury table}
kable(select(topFatalities, event, fatalities.per.year),digits=2)
kable(select(topInjuries, event, injuries.per.year),digits=2)
```

Comparing the above tables, several entries are found in both of them.

## Economic consequences
After evaluating the danger to population, also the cost factor will be investigated. The costs sum up from crop and property damage. First the total sum of all costs for each event is calculated.
```{r cost preparation}
costs <- with(data, tapply(costs, evtype, sum))
costdata <- data.frame(cost=as.vector(costs), event=as.factor(names(amounts)))

costdata[sample(nrow(costdata), 20),]
```

Next events without costs are dismissed.
```{r dismiss zeros costs}
#healthdata <- filter(healthdata, amount>0)
```

This plot depictures the money spent for each event within the recorded `r print(2011-1950)` years
```{r cost plot}
p <- ggplot(costdata, aes(x=reorder(event, desc(event)),y=cost))
p + geom_bar(stat='identity', position='dodge') + coord_flip() + labs(x="event types", y="costs [USD]")
```

The graphic again indicates several costly. The 10 most expensive events are found and divided by the number of years recorded to calculate the impact per year
```{r top costs}
topCosts <- head(arrange(costdata, desc(cost)),10)
topCosts <- mutate(topCosts, cost.per.year=cost/(2011-1950))
```

The following table notes down the most 10 expensive events in USD per year:
```{r cost table}
kable(select(topCosts, event, cost.per.year),digits=2)
```

This evaluation shows about the same events as the human health evaulation.

** 1 - 3 figures with plot have descriptive caption **
** question: Across the united states which types of events are most harmful to population health? **
** question: Across the united states which types of events have the greatest economic consequences? **
** reproducible results, summaries, figures, tables **
** document is written as it might be read by government or municipal manager **
** synopsis **

## 